---
title: "Exploring and processing metabarcoding data with metagMisc"
author: "Vladimir Mikryukov"
date: "`r Sys.Date()`"
output: github_document
vignette: >
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This vignette demonstrates a typical metabarcoding data analysis workflow using the `metagMisc` package. 
We'll walk through the process of exploring, filtering, and analyzing amplicon sequencing data, 
from initial data inspection to comparative analysis between sample groups.

The `metagMisc` package contains multiple functions for working with phyloseq objects 
([McMurdie and Holmes 2013](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0061217)), 
which provide a convenient interface and data structure for working with metabarcoding/microbiome data in R. 

> [!NOTE]  
> In the output, columns might be named "OTU" (Operational Taxonomic Unit), 
> but this corresponds to whatever units were used in the original dataset (e.g., ASVs, genes, species, etc.) 
> and should be treated generically.

## Load packages and data

```{r load-packages}
library(phyloseq)
library(ggplot2)
library(metagMisc)
```

We'll use the `GlobalPatterns` dataset from the phyloseq package, 
which contains 16S rRNA amplicons from various environmental samples obtained by 
[Caporaso et al. (2011)](https://www.pnas.org/doi/full/10.1073/pnas.1000080107).

```{r load-data}
data(GlobalPatterns)
GlobalPatterns
```

## 1. Initial data exploration

### Dataset overview

First, let's get a comprehensive overview of our dataset:

```{r data-summary}
phyloseq_summary(GlobalPatterns, cols = "GlobalPatterns")
```

### Taxonomic annotation quality

One of the first things to check is how well the sequences are taxonomically annotated:

```{r taxonomic-resolution, eval = TRUE}
phyloseq_taxonomic_resolution(GlobalPatterns)
```

This plot shows the percentage of OTUs and reads annotated at each taxonomic rank. 
The higher the resolution, the better the taxonomic annotation quality.

### Taxonomic diversity across sample types

Let's examine the taxonomic diversity across different sample types. First, we'll combine samples by type and count taxa at different ranks:

```{r taxa-by-sample-type}
# Combine samples by sample type
gcomb <- phyloseq_combine_samples(GlobalPatterns, variable = "SampleType")
phyloseq_ntaxlevels(gcomb, format = "wide")
```

We can also visualize this information:

```{r taxa-diversity-plot}
# Create plot-friendly table
gtab <- phyloseq_ntaxlevels(gcomb, format = "long")
gtab$TaxRank <- factor(gtab$TaxRank, levels = rank_names(gcomb))
gtab <- gtab[!gtab$TaxRank %in% "Kingdom", ]

ggplot(data = gtab, aes(y = Sample, x = N.tax.levels, fill = Sample)) +
    geom_bar(stat = "identity") +
    facet_wrap(~ TaxRank, scales = "free") +
    labs(y = "Sample type", x = "Number of unique taxa") +
    theme(legend.position = "none")
```

```
