---
title: "Exploring and processing metabarcoding data with metagMisc"
author: "Vladimir Mikryukov"
date: "`r Sys.Date()`"
output: github_document
vignette: >
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This vignette demonstrates a typical metabarcoding data analysis workflow using the `metagMisc` package. 
We'll walk through the process of exploring, filtering, and analyzing amplicon sequencing data, 
from initial data inspection to comparative analysis between sample groups.

The `metagMisc` package contains multiple functions for working with phyloseq objects 
([McMurdie and Holmes 2013](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0061217)), 
which provide a convenient interface and data structure for working with metabarcoding/microbiome data in R. 


## Load packages and data

```{r load-packages}
library(phyloseq)
library(ggplot2)
library(metagMisc)
```
