---
title: "Exploring and processing metabarcoding data with metagMisc"
author: "Vladimir Mikryukov"
date: "`r Sys.Date()`"
output: github_document
vignette: >
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This vignette demonstrates a typical metabarcoding data analysis workflow using the `metagMisc` package. 
We'll walk through the process of exploring, filtering, and analyzing amplicon sequencing data, 
from initial data inspection to comparative analysis between sample groups.

The `metagMisc` package contains multiple functions for working with phyloseq objects 
([McMurdie and Holmes 2013](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0061217)), 
which provide a convenient interface and data structure for working with metabarcoding/microbiome data in R. 

> [!NOTE]  
> In the output, columns might be named "OTU" (Operational Taxonomic Unit), 
> but this corresponds to whatever units were used in the original dataset (e.g., ASVs, genes, species, etc.) 
> and should be treated generically.

## Load packages and data

```{r load-packages}
library(phyloseq)
library(ggplot2)
library(metagMisc)
```

We'll use the `GlobalPatterns` dataset from the phyloseq package, 
which contains 16S rRNA amplicons from various environmental samples obtained by 
[Caporaso et al. (2011)](https://www.pnas.org/doi/full/10.1073/pnas.1000080107).

```{r load-data}
data(GlobalPatterns)
GlobalPatterns
```

## 1. Initial data exploration

### Dataset overview

First, let's get a comprehensive overview of our dataset:

```{r data-summary}
phyloseq_summary(GlobalPatterns, cols = "GlobalPatterns")
```

### Taxonomic annotation quality

One of the first things to check is how well the sequences are taxonomically annotated:

```{r taxonomic-resolution, eval = TRUE}
phyloseq_taxonomic_resolution(GlobalPatterns)
```

This plot shows the percentage of OTUs and reads annotated at each taxonomic rank. 
The higher the resolution, the better the taxonomic annotation quality.

### Taxonomic diversity across sample types

Let's examine the taxonomic diversity across different sample types. First, we'll combine samples by type and count taxa at different ranks:

```{r taxa-by-sample-type}
# Combine samples by sample type
gcomb <- phyloseq_combine_samples(GlobalPatterns, variable = "SampleType")
phyloseq_ntaxlevels(gcomb, format = "wide")
```

We can also visualize this information:

```{r taxa-diversity-plot}
# Create plot-friendly table
gtab <- phyloseq_ntaxlevels(gcomb, format = "long")
gtab$TaxRank <- factor(gtab$TaxRank, levels = rank_names(gcomb))
gtab <- gtab[!gtab$TaxRank %in% "Kingdom", ]

ggplot(data = gtab, aes(y = Sample, x = N.tax.levels, fill = Sample)) +
    geom_bar(stat = "identity") +
    facet_wrap(~ TaxRank, scales = "free") +
    labs(y = "Sample type", x = "Number of unique taxa") +
    theme(legend.position = "none")
```

## 2. Data filtering and quality control

Real metabarcoding datasets often contain many rare taxa that may represent sequencing errors or contaminants. Let's apply several filtering strategies to clean our data.

### Prevalence-based filtering

First, let's examine the prevalence patterns of the most abundant phyla:

```{r prevalence-plot}
# Subset to major phyla for clearer visualization
GP_major <- subset_taxa(GlobalPatterns,
  Phylum %in% c("Acidobacteria", "Actinobacteria", "Firmicutes", "Verrucomicrobia"))

phyloseq_prevalence_plot(GP_major, taxcolor = "Phylum",
  facet = TRUE, point_alpha = 0.5, prev.trh = 0.05)
```

The horizontal line shows the 5% prevalence threshold. OTUs above this line are present in at least 5% of samples.

Remove OTUs that are not present in at least 5% of samples:

```{r prevalence-filter}
# Apply prevalence filter
GP_prev <- phyloseq_filter_prevalence(GlobalPatterns, prev.trh = 0.05)
```

Summary of the filtered dataset:

```{r summary-filtered_demo, eval = TRUE}
phyloseq_summary(
  GlobalPatterns, GP_prev,
  cols = c("Before filtering", "After filtering"),
  more_stats = TRUE)
```

```{r summary-filtered, echo=FALSE, message=FALSE, warning=FALSE, eval = FALSE}
# Use nice printing method (does not work well with knitr)
metagMisc:::format.phyloseq_summary_wide(
  phyloseq_summary(
    GlobalPatterns, GP_prev,
    cols = c("Before filtering", "After filtering"),
    more_stats = TRUE),
  use_color = FALSE)
```

### Abundance-based filtering

Remove very rare taxa based on relative abundance:

```{r abundance-filter}
# Remove taxa with mean relative abundance < 0.01%
GP_abund <- phyloseq_filter_taxa_rel_abund(GP_prev, frac = 0.0001)

# Compare datasets
phyloseq_summary(
  GlobalPatterns, GP_prev, GP_abund, 
  cols = c("Original", "Prevalence_filtered", "Abundance_filtered"))
```

### Total abundance fraction filtering

Remove taxa that contribute less than a specified fraction of the total abundance across the dataset:

```{r total-fraction-filter}
# Retain OTUs with at least 0.1% of total abundance
GP_totfrac <- phyloseq_filter_taxa_tot_fraction(GP_prev, frac = 0.001)

# Compare
phyloseq_summary(
  GlobalPatterns, GP_prev, GP_totfrac,
  cols = c("Original", "Prevalence_filtered", "TotalFrac_filtered"))
```

### Retain most abundant taxa

Keep only the most abundant taxa using either a percentage or a fixed number:

```{r top-taxa-filter}
# Top 5% most abundant taxa
GP_top <- phyloseq_filter_top_taxa(GlobalPatterns, perc = 5)

# Alternatively: retain a fixed number of top taxa
GP_top_n <- phyloseq_filter_top_taxa(GlobalPatterns, n = 500)

phyloseq_summary(
  GlobalPatterns, GP_top, GP_top_n,
  cols = c("Original", "TopPct_5", "TopN_500"))
```

You can also explore how different percentage thresholds affect retained abundance per sample:

```{r top-taxa-range}
phyloseq_filter_top_taxa_range(GlobalPatterns, show_plot = TRUE)
```

### Sample richness filter

Optionally remove samples with low richness (few observed taxa):

```{r richness-filter}
GP_rich <- phyloseq_richness_filter(GlobalPatterns, mintaxa = 3000)
phyloseq_summary(GP_rich, cols = "Richness_filtered")
```

### Sample-wise abundance trimming

Remove low-abundance OTUs within each sample (sample-wise filtering):

```{r sample-wise-filter}
# Remove OTUs with <10 reads in each sample
GP_trim <- phyloseq_filter_sample_wise_abund_trim(GlobalPatterns, minabund = 10)

# Final comparison
phyloseq_summary(
  GlobalPatterns, GP_trim, 
  cols = c("Original", "Filtered"),
  more_stats = TRUE)
```

```
